---
title: "ATL08_Derived_Growth_Using_HT"
format: html
editor: visual
---

## Libraries

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(arrow)
library(dplyr)
library(mase)
```

## Reading Data

Our working set is ICESat-2 ATL08 segments with:

-   \> 3 years recovery

-   disturbance before the year of collection

Further, we remove segments collected in 2019 which demonstrated spatial differences in forest height from equivalence testing

```{r}
workingSet <- read_parquet(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\segments\workingSetPolygonsWITHALSMetrics.parquet)") |> 
  filter(!(year == 2019))
```

## Net Growth
```{r}
y20 <- workingSet |> filter(year == 2020) |> pull(h_canopy)
ht20 <- horvitzThompson(y = y20,
                        N = length(y20), 
                        var_est = FALSE)

y24 <- workingSet |> filter(year == 2024) |> pull(h_canopy)
ht24 <- horvitzThompson(y = y24,
                        N = length(y24),
                        var_est = FALSE)


# Calculating difference of 2024 and 2020 mean segment heights as proxy for net growth 
atl08_net_growth <- ht24$pop_mean - ht20$pop_mean
atl08_net_growth
```

## Calculating variances
Doing this by hand, separately for t1 and t2 heights
```{r}
# helper function for calculating variance from one of our samples
sample_variance <- function(sample) {
  # the sum of the squared difference for each element in the sample, divided it by (n-1)
  sum((sample - mean(sample))^2) / (length(sample) - 1)
}

var20 <- sample_variance(y20)

var24 <- sample_variance(y24)
```

## Pooled variance 
For ATL08, this is t1 variance plus t2 variance
```{r}
atl08_pooled_var <- var20 + var24
atl08_pooled_var
```

## Confidence interval
```{r}
atl08_confidence_interval <- 1.96 * sqrt(atl08_pooled_var)
atl08_confidence_interval
```
