---
title: "ATL08_Derived_Growth_Using_HT"
format: html
editor: visual
---

## Libraries

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(RobustLinearReg)
library(arrow)
library(dplyr)
library(glue)
library(mase)
library(patchwork)
```

## Global Variables

```{r}
# Number of iterations for bootstrapping (only used for Theil-Sen regression)
numberOfIterations <- 10000

# Levels of our factors
nlcd_types <- c("Deciduous Forest", "Evergreen Forest", "Mixed Forest", "Woody Wetlands")
disturbance_presence_types <- c("Yes", "No")
fast_loss_groups <- c("<10 Years", "10-20 Years", "20-30 Years", "30+ Years")

# Custom plotting colors for each factor level
custom_colors <- c(
  # forest cover 
  "Deciduous Forest" = "#a1d99b",
  "Evergreen Forest" = "#74c476",
  "Mixed Forest" = "#31a354",
  "Woody Wetlands" = "#006d2c",
  
  # disturbance binary (reds)
  "Yes" = "#fb6a4a",
  "No" = "#cb181d",
  
  # years since
  "<10 Years" = "#9ecae1",
  "10-20 Years" = "#6baed6",
  "20-30 Years" = "#3182bd",
  "30+ Years" = "#08519c"
)

customTheme <-theme(
      plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
      axis.title = element_text(size = 20),
      axis.line = element_line(linewidth = 0.5, color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.text = element_text(size=15),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "none",
      coord_cartesian(clip = "off")
    )
```

## Reading Data

Our working set is ICESat-2 ATL08 segments with:

-   \> 3 years recovery

-   disturbance before the year of collection

Further, we remove segments collected in 2019 which demonstrated spatial differences in forest height from equivalence testing

```{r}
workingSet <- read_parquet(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\segments\workingSetPolygonsWITHALSMetrics.parquet)") |> 
  filter(!(year == 2019))
```

## Net Growth

### Un-stratified Horvitz-Thompson results

```{r}
y20 <- workingSet |> filter(year == 2020) |> pull(h_canopy)
ht20 <- horvitzThompson(y = y20,
                        N = length(y20), 
                        var_est = TRUE,
                        var_method = "LinHH")

y24 <- workingSet |> filter(year == 2024) |> pull(h_canopy)
ht24 <- horvitzThompson(y = y24,
                        N = length(y24),
                        var_est = TRUE,
                        var_method = "LinHH")


unstratified_atl08_ht <- data.frame(
     first_year_length = length(y20),
     first_year_pop_mean = ht20$pop_mean,
     first_year_var = ht20$pop_mean_var,
     last_year_length = length(y24),
     last_year_pop_mean = ht24$pop_mean,
     last_year_var = ht24$pop_mean_var,
     net_growth = ht24$pop_mean - ht20$pop_mean
  )
unstratified_atl08_ht
```
### Confidence interval
```{r}
atl08_t1_var <- var(y20)
atl08_t2_var <- var(y24)
# atl08_t1_var <- unstratified_atl08_ht$first_year_var
# atl08_t2_var <- unstratified_atl08_ht$last_year_var
atl08_pooled_var <- atl08_t1_var + atl08_t2_var
atl08_nft <- 1.96 * sqrt(atl08_pooled_var) 
atl08_nft
```

#### Graphing
```{r}
atl08_start_heights_plot <- ggplot(mapping = aes(y20)) + 
  geom_histogram(binwidth = 1, fill = "#3689D2", color = "black") +
  scale_x_continuous(breaks = seq(0, 40, by = 5)) +
  geom_vline(xintercept=mean(y20), color='red', lwd=.75)+
  annotate("text", x = 14, y = 275, size=7, face='bold', color='black', label = glue("Mean: {round(mean(y20), 2)}"))+
  labs(title = "Start ATL08 Heights",
       x = "Height (m)",
       y = "Frequency")+
  theme_minimal() +
  customTheme

atl08_start_heights_plot

atl08_end_heights_plot <- ggplot(mapping = aes(y24)) + 
  geom_histogram(binwidth = 1, fill = "#8A7D63", color = "black") +
  scale_x_continuous(breaks = seq(0, 40, by = 5)) + 
  geom_vline(xintercept=mean(y24), color='red', lwd=.75)+
  annotate("text", x = 15, y = 775, size=7, face='bold', color='black', label = glue("Mean: {round(mean(y24), 2)}"))+
  labs(title = "End ATL08 Heights",
       x = "Height (m)",
       y = "Frequency")+
  theme_minimal() +
  customTheme 

atl08_end_heights_plot

combined_atl08_heights_plot <- (atl08_start_heights_plot + atl08_end_heights_plot)

ggsave(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\figures\combined_atl08_heights_plot.png)", plot = combined_atl08_heights_plot, bg = "transparent",
width = 16, height = 8, units = "in", dpi = 300)

combined_atl08_heights_plot
```

### Horvitz-Thompson estimator function

```{r}
HTByFactor <- function(dataframe, factorName, factorLevel, heightsColumn, firstYear, lastYear){
  
  print(glue("Running HT estimators for {factorName} {factorLevel}, for years {firstYear} and {lastYear}"))
  
  # Pulling heights from our first year of data
  first_year_heights <- dataframe |> 
    filter(year == firstYear) |> 
    filter(.data[[factorName]] == factorLevel) |> 
    pull(.data[[heightsColumn]])
  
  # getting the length (N) of the first year's population
  first_year_length <- length(first_year_heights)
  
  # running HT estimator on the first year's population
  first_year_ht <- horvitzThompson(y = first_year_heights, 
                          N = first_year_length,
                          var_est = FALSE)
  
  # calculating variance of first year's canopy heights
  first_year_var <- var(first_year_heights)
  
  
  # Pulling heights from our last year of data
  last_year_heights <- dataframe |> 
    filter(year == lastYear) |> 
    filter(.data[[factorName]] == factorLevel) |> 
    pull(.data[[heightsColumn]])
  
  # getting the length (N) of the last year's population
  last_year_length <- length(last_year_heights)
  
  # running HT estimator on the last year's population
  last_year_ht <- horvitzThompson(y = last_year_heights,
                                  N = last_year_length, 
                                  var_est = FALSE)
  
  # calculating variance of last year's canopy heights
  last_year_var <- var(last_year_heights)
  
  # difference in first year and last year canopy height as proxy for net growth
  diff_means <-  last_year_ht$pop_mean - first_year_ht$pop_mean 
  
  # confidence interval of the difference in means
  diff_ci <- 1.96 * sqrt(first_year_var + last_year_var)

  # combining results into a dataframe
   factor_ht_results <- data.frame(
     factor = factorName,
     level = factorLevel,
     first_year_n = first_year_length,
     last_year_n = last_year_length,
     first_year_pop_mean = first_year_ht$pop_mean, # mean canopy height at the start of the segment timeframe
     last_year_pop_mean = last_year_ht$pop_mean, # mean canopy height at the end of the segment timeframe
     diff_in_means = diff_means, 
     CI_of_diff = diff_ci
   )

  cat("\n")

  return(factor_ht_results)
}

# wrapper function to run a HT estimator for each level of the supplied factor
run_HT_for_factor <- function(levels, fac, dataframe, heightsColumn, firstYear, lastYear) {
  results <- lapply(levels, function(lvl) {
    res <- HTByFactor(
      dataframe = dataframe,
      factorName = fac,
      factorLevel = lvl,
      heightsColumn = heightsColumn,
      firstYear = firstYear,
      lastYear = lastYear
    )
    return(res)
  }) |> bind_rows()
  return(results)
}
```

### Stratified Horvitz-Thompson estimator results

```{r}
nlcd_ht_results <- run_HT_for_factor(nlcd_types, "NLCD_2023_Label", workingSet, "h_canopy", 2020, 2024)

disturbance_ht_results <- run_HT_for_factor(disturbance_presence_types, "DisturbancePresence", workingSet, "h_canopy", 2020, 2024)

fast_loss_ht_results <- run_HT_for_factor(fast_loss_groups, "FastLossAgeGroup", workingSet, "h_canopy", 2020, 2024)

# putting them all together
final_ht_results <- bind_rows(nlcd_ht_results, disturbance_ht_results, fast_loss_ht_results) %>%
  mutate(
    factor = recode(factor,
      "NLCD_2023_Label"   = "Forest Cover Type",
      "DisturbancePresence" = "Disturbance Presence",
      "FastLossAgeGroup" = "Years Since Disturbance"
    )
  )

write.csv(final_ht_results, r"(C:\Users\poncy\Desktop\Masters\Research\Publication\R\final_ht_results.csv)")

final_ht_results
```

### Graphing

```{r}
nlcd_ht_plot <- final_ht_results |> 
  filter(factorName == "Forest Cover Type") |> 
  # mutate(level = factor(level, levels = levels_order)) |> 
  ggplot(aes(x = level, y = diff_in_means, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
    # labeling bars
    geom_text(
      aes(label = round(diff_in_means, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Forest Cover Type",
      y = "Net Growth (m)",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme

disturbance_ht_plot <- final_ht_results |> 
  filter(factorName == "Disturbance Presence") |> 
  # mutate(level = factor(level, levels = levels_order)) |> 
  ggplot(aes(x = level, y = diff_in_means, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
     # labeling bars
    geom_text(
      aes(label = round(diff_in_means, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Disturbance Presence",
      y = "Net Growth (m)",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme +
  theme(axis.text.y = element_blank(),axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank()) # removing y-axis and labels

years_since_ht_plot <- final_ht_results |> 
  filter(factorName == "Years Since Disturbance") |> 
  # mutate(level = factor(level, levels = levels_order)) |> 
  ggplot(aes(x = level, y = diff_in_means, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
    # labeling bars
    geom_text(
      aes(label = round(diff_in_means, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Years Since Disturbance",
      y = "Net Growth (m)",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme +
  theme(axis.text.y = element_blank(),axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank()) # removing y-axis and labels

combined_ht_plot <- (nlcd_ht_plot + disturbance_ht_plot + years_since_ht_plot ) & scale_y_continuous(limits = c(0,3.5))

ggsave(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\figures\combined_ht_plot.png)", plot = combined_ht_plot, bg = "transparent",
width = 16, height = 8, units = "in", dpi = 300)

combined_ht_plot
```

## Incremental Growth

### Graphing

#### Slope plot

```{r}
final_tsr_results <- read_csv(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\R\final_tsr_results.csv)")

nlcd_slope_plot <- final_tsr_results |> 
  filter(factorName == "Forest Cover Type") |> 
  # mutate(level = factor(level, levels = levels_order)) |> 
  ggplot(aes(x = level, y = median_slope, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
    # only labeling significant bars
    geom_text(
      data = \(d) d |> filter(median_p_value < 0.05),
      aes(label = round(median_slope, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Forest Cover Type",
      y = "Slope (meters per year)",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme 

disturbance_slope_plot <- final_tsr_results |> 
  filter(factorName == "Disturbance Presence") |> 
  # mutate(level = factor(level, levels = levels_order)) |> 
  ggplot(aes(x = level, y = median_slope, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
    # only labeling significant bars
    geom_text(
      data = \(d) d |> filter(median_p_value < 0.05),
      aes(label = round(median_slope, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Disturbance Presence",
      y = "Slope (meters per year))",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme + 
    theme(axis.text.y = element_blank(),axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())

years_since_slope_plot <- final_tsr_results |> 
  filter(factorName == "Years Since Disturbance") |> 
  # mutate(level = factor(level, levels = levels_order)) |> 
  ggplot(aes(x = level, y = median_slope, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
    # only labeling significant bars
    geom_text(
      data = \(d) d |> filter(median_p_value < 0.05),
      aes(label = round(median_slope, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Years Since Disturbance",
      y = "Slope (meters per year)",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme + 
    theme(axis.text.y = element_blank(),axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())


# Combine with patchwork and add a single overarching title
combined_slope_plot <- (nlcd_slope_plot + disturbance_slope_plot + years_since_slope_plot ) & scale_y_continuous(limits = c(0, 0.65))


ggsave(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\figures\TSR_Slope_Plot.png)", plot = combined_slope_plot, bg = "transparent",
width = 16, height = 8, units = "in", dpi = 300)

combined_slope_plot

```

#### Intercept plot

```{r}
nlcd_intercept_plot <- final_tsr_results |> 
  filter(factorName == "Forest Cover Type") |> 
  ggplot(aes(x = level, y = median_intercept, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
    # only labeling significant bars
    geom_text(
      data = \(d) d |> filter(median_p_value < 0.05),
      aes(label = round(median_intercept, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Forest Cover Type",
      y = "Baseline (meters)",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme 

disturbance_intercept_plot <- final_tsr_results |> 
  filter(factorName == "Disturbance Presence") |> 
  # mutate(level = factor(level, levels = levels_order)) |> 
  ggplot(aes(x = level, y = median_intercept, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
    # only labeling significant bars
    geom_text(
      data = \(d) d |> filter(median_p_value < 0.05),
      aes(label = round(median_intercept, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Disturbance Presence",
      y = "Baseline (meters)",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme + 
    theme(axis.text.y = element_blank(),axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())

years_since_intercept_plot <- final_tsr_results |> 
  filter(factorName == "Years Since Disturbance") |> 
  # mutate(level = factor(level, levels = levels_order)) |> 
  ggplot(aes(x = level, y = median_intercept, fill = level)) +
    geom_col(width = 0.8) +
    scale_fill_manual(values = custom_colors) +
    # only labeling significant bars
    geom_text(
      data = \(d) d |> filter(median_p_value < 0.05),
      aes(label = round(median_intercept, 2)),
      vjust = -0.5,
      size = 6,
      fontface = "bold",
      show.legend = FALSE
    ) +
    labs(
      x = "Years Since Disturbance",
      y = "Baseline (meters)",
      title = element_blank()
    ) +
    theme_minimal() +
    customTheme + 
    theme(axis.text.y = element_blank(),axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank())


# Combine with patchwork and add a single overarching title
combined_intercept_plot <- (nlcd_intercept_plot + disturbance_intercept_plot + years_since_intercept_plot) & scale_y_continuous(limits = c(0, 25))


ggsave(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\figures\TSR_Intercept_Plot.png)", plot = combined_intercept_plot, bg = "transparent",
width = 16, height = 8, units = "in", dpi = 300)

combined_intercept_plot
```
