---
title: "NFI_Reference_Growth"
format: html
editor: visual
---

## Importing packages & setting path to local copy of North Carolina FIADB

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(mase)
library(dplyr)
library(RobustLinearReg)
library(ggplot2)
library(glue)
library(patchwork)
data_path = file.path(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\FIAData\FIADB_Direct\Working)")
source(file.path(file.path(data_path,'FIADB_Direct_source_all.R')))
```

## Global Variables

```{r}
# The number of years in out ATL08 timeframe (4 years)
timeframeLength <- 4

customTheme <-theme(
      plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
      axis.title = element_text(size = 20),
      axis.line = element_line(linewidth = 0.5, color = "black"),
      axis.ticks = element_line(color = "black"),
      axis.text = element_text(size=15),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "none",
      coord_cartesian(clip = "off")
    )

# Global minima of 2 meters
globalMinima <- 2

# Global maxima of 39.423 meters
globalMaxima <- 39.423476440429695
```

## Extracting NFI plot data from FIADB

```{r}
NC_plots_AGB <- PLOT_obs(EVAL_GRP = 372023, ATTRIBUTE_NBR = 10,GRP_BY_ATTRIB = c('unitcd','countycd')) |> 
  filter(COUNTYCD %in% c(13, 15, 17, 19, 29, 31, 41, 47, 49, 53, 55, 61, 65, 69, 73, 79, 83, 91, 95,101, 103, 107, 117, 127, 129, 131, 133, 137, 139, 141, 143, 147, 155, 163, 177, 181, 185, 187, 191, 195)
  ) # here i'm generating plot-level estimates from all the plots in evaluation group 372023, corresponding to North Carolina plots visited in 2023. Then, filtering to just those within the counties of the study area using the county code

# unique plot codes for EVAL_GRP 372023 within study area
PLT_CN <- unique(NC_plots_AGB$PLT_CN)

# this grabs plots from the PLOT table with sequence numbers matching our NC_plots_AGB 
plot <- GET_record(TABLE_NAME = 'plot',VAR_NAME = 'cn',VAR_VALUES = PLT_CN) |> 
  rename(PLT_CN = CN)

# this grabs conditions from the CONDITION table with sequnce numbers matching our NC_plots_AGB 
cond <- GET_record(TABLE_NAME = 'cond',VAR_NAME = 'plt_cn',VAR_VALUES = PLT_CN)

# unique plot conditions, as there are multiple plot sequences in cond table
PLT_CN_ALL <- unique(cond$PLT_CN)

# this only gets plots which match the unique sequence numbers
plot <- plot |> filter(PLT_CN %in% PLT_CN_ALL)

# getting individual tree records
tree <- GET_record(TABLE_NAME = 'tree',VAR_NAME = 'plt_cn',VAR_VALUES = PLT_CN_ALL) |> 
  rename(TRE_CN = CN)


### Removing dates
plot <- plot[,-grep("_DATE",names(plot))]
cond <- cond[,-grep("_DATE",names(cond))]
tree <- tree[,-grep("_DATE",names(tree))]


### Joining plot, cond, tree tables
plot_cond <- left_join(plot,cond)
plot_cond_tree <- left_join(plot_cond,tree)
# test <- anti_join(tree,plot_cond_tree) # Trees from plots with partial non-SYP fortypecd  

```

## Current NFI plot measurements

```{r}
### Creating current_tree dataframe for current measurements of individual trees
current_tree <- plot_cond_tree |> 
  dplyr::select(PLT_CN,PREV_PLT_CN,STATECD,UNITCD,COUNTYCD,MEASYEAR,REMPER,CONDID,COND_STATUS_CD,FORTYPCD,STDORGCD,
                STDORGSP,DSTRBCD1,DSTRBYR1, TRTCD1,TRE_CN,PREV_TRE_CN,SUBP,TREE,STATUSCD,SPCD,DIA,HT,ACTUALHT,CCLCD,
                DAMLOC1,DAMTYP1,DAMSEV1) |> 
  rename(PLT_CN2 = PLT_CN,
         MEASYEAR2 = MEASYEAR,
         CONDID2 = CONDID,
         COND_STATUS_CD2 = COND_STATUS_CD,
         FORTYPCD2 = FORTYPCD,
         STDORGCD2 = STDORGCD,
         STDORGSP2 = STDORGSP,
         DSTRBCD12 = DSTRBCD1,
         DSTRBYR12 = DSTRBYR1,
         TRTCD12 = TRTCD1,
         TRE_CN2 = TRE_CN,
         STATUSCD2 = STATUSCD,
         DIA2 = DIA, HT2 = HT,
         ACTUALHT2 = ACTUALHT,
         CCLCD2 = CCLCD,
         DAMLOC12 = DAMLOC1,
         DAMTYP12 = DAMTYP1,
         DAMSEV12 = DAMSEV1)
```

## Previous NFI Plot Measurements

```{r}
# grabs the previous sequence numbers for each plot
PREV_PLT_CN <- unique(plot$PREV_PLT_CN) 

# then gets the plot records for those previous plot sequence numbers
prev_plot <- GET_record('plot','cn',PREV_PLT_CN) |> 
  rename(PLT_CN = CN)

# then grabs the conditions for the previous plot records
prev_cond <- GET_record(TABLE_NAME = 'cond',VAR_NAME = 'plt_cn',VAR_VALUES = PREV_PLT_CN) |> 
  filter(FORTYPCD %in% c(141, 142, 161, 162, 166, 171, 402, 403, 404, 406, 407, 409, 501, 503, 504, 505, 506, 507, 508, 511, 512, 514, 515, 516, 517, 519, 520, 962, 995, 601, 602, 605, 606, 607, 608, 609, 701, 702, 703, 704, 705, 706, 708, 709))

# finally, joining tree data for previous plot records
prev_tree <- GET_record(TABLE_NAME = 'tree',VAR_NAME = 'plt_cn',VAR_VALUES = PREV_PLT_CN) |> 
  rename(TRE_CN = CN)

# drop _DATE variables from previous plot records
prev_plot <- prev_plot[,-grep("_DATE",names(prev_plot))]
prev_cond <- prev_cond[,-grep("_DATE",names(prev_cond))]
prev_tree <- prev_tree[,-grep("_DATE",names(prev_tree))]


### Joining previous plot, cond, tree

# joinging condiiton and tree data from previous plot records
prev_plot_cond <- left_join(prev_plot,prev_cond)
prev_plot_cond_tree <- left_join(prev_plot_cond,prev_tree)

# establishing a dataframe of previous tree measurments
previous_tree <- prev_plot_cond_tree |> 
  dplyr::select(PLT_CN,STATECD,UNITCD,COUNTYCD,MEASYEAR,CONDID,COND_STATUS_CD,FORTYPCD,STDORGCD,
                STDORGSP,DSTRBCD1,DSTRBYR1, TRTCD1,TRE_CN,SUBP,TREE,STATUSCD,SPCD,DIA,HT,ACTUALHT,CCLCD,
                DAMLOC1,DAMTYP1,DAMSEV1)
```

## Joining current and previous NFI plot measurements

```{r}
nfi_ht_df <- left_join(current_tree,previous_tree,
                  by=join_by(STATECD,UNITCD,COUNTYCD,SUBP,TREE,SPCD,
                             PREV_PLT_CN == PLT_CN,PREV_TRE_CN == TRE_CN)) |> 
  # Getting measurements in meters
  mutate(
    HT_m = HT * .3048,
    HT2_m = HT2 * .3048,
    ACTUALHT_m = ACTUALHT * .3048,
    ACTUALHT2_m = ACTUALHT2 * .3048
  ) |> 
  filter(STATUSCD <= 1 & STATUSCD2 <= 1) # STATUS CODE to keep alive trees

# -------------- Mapping forest type ----------------

fortypMapping <- c(
  # Evergreens
  "141" = "Evergreen Forest", "142" = "Evergreen Forest", "161" = "Evergreen Forest",
  "162" = "Evergreen Forest", "166" = "Evergreen Forest", "171" = "Evergreen Forest",
  "606" = "Evergreen Forest",
  # then deciduous
  "501" = "Deciduous Forest", "503" = "Deciduous Forest", "504" = "Deciduous Forest",
  "505" = "Deciduous Forest", "506" = "Deciduous Forest", "507" = "Deciduous Forest",
  "508" = "Deciduous Forest", "511" = "Deciduous Forest", "512" = "Deciduous Forest",
  "514" = "Deciduous Forest", "515" = "Deciduous Forest", "516" = "Deciduous Forest",
  "517" = "Deciduous Forest", "519" = "Deciduous Forest", "520" = "Deciduous Forest",
  "701" = "Deciduous Forest", "702" = "Deciduous Forest", "703" = "Deciduous Forest",
  "704" = "Deciduous Forest", "705" = "Deciduous Forest", "706" = "Deciduous Forest",
  "708" = "Deciduous Forest", "709" = "Deciduous Forest", "962" = "Deciduous Forest",
  # then mixed
  "402" = "Mixed Forest", "403" = "Mixed Forest", "404" = "Mixed Forest", "406" = "Mixed Forest",
  "407" = "Mixed Forest", "409" = "Mixed Forest",
  # then woody wetlands
  "601" = "Woody Wetlands", "602" = "Woody Wetlands", "605" = "Woody Wetlands",
  "607" = "Woody Wetlands", "608" = "Woody Wetlands", "609" = "Woody Wetlands",
  # i'll leave unknown just as-is
  "999" = "Unknown"
)

# then applying the mapping
nfi_ht_df <- nfi_ht_df |>
  mutate(NLCD_2023_Label = recode(as.character(FORTYPCD), !!!fortypMapping)
         )

# -------------- Mapping disturbance presence ----------------

# encoding disturbance presence
nfi_ht_df$DisturbancePresence <- ifelse(nfi_ht_df$DSTRBCD12 == 0, 'No', 'Yes')

# adding years since disturbance
nfi_ht_df$YearsSinceDisturbance <- nfi_ht_df$MEASYEAR2 - nfi_ht_df$DSTRBYR12 

# here marking continuous disturbances (9999 for disturbance year) and were negative in prev calculation
nfi_ht_df$YearsSinceDisturbance <- ifelse(nfi_ht_df$YearsSinceDisturbance < 0, 'Continuous', nfi_ht_df$YearsSinceDisturbance)
```

## Filtering NFI observations

```{r}
nfi_df_filtered <- nfi_ht_df |> 
  filter(HT2_m >= HT_m) |> # removing negative changes in total height 
  filter(ACTUALHT2_m >= ACTUALHT_m) |> # removing negative changes in actual height
  
  filter(HT_m >= globalMinima & HT_m <= globalMaxima) |> # filtering previous total heights to within our min and max values
  filter(HT2_m >= globalMinima & HT2_m <= globalMaxima) |>  # filtering current total heights to within our min and max values
  
  filter(ACTUALHT_m >= globalMinima & ACTUALHT_m <= globalMaxima) |>  # filtering previous actual heights to within our min and max values
  filter(ACTUALHT2_m >= globalMinima & ACTUALHT2_m <= globalMaxima) |>  # filtering current actual heights to within our min and max values
  
  filter(REMPER >= (timeframeLength - 0.1) & REMPER <= (timeframeLength + 1.1)) # LIMITING REMEASUREMENT PERIOD TO MATCH ATL08 TIMEFRAME
```

## Calculating plot-level net change in tree height and tree height increment

Here we aggregate individual-tree observations up to the plot level mean.

```{r}
# Mean plot-level tree height at time 1
mean_height_t1 <- nfi_df_filtered |> 
  group_by(PLT_CN2) |> # grouping by plot sequence numbers at revisit
  summarise("Time 1" = mean(ACTUALHT_m, na.rm = TRUE)) # plot-level mean actual tree height at t1

# Mean plot-level tree height at time 2
mean_height_t2 <- nfi_df_filtered |> 
  group_by(PLT_CN2) |> # grouping by plot sequence numbers at revisit
  summarise("Time 2" = mean(ACTUALHT2_m, na.rm = TRUE)) # plot-level mean actual tree height at t2

# merging net change and increment into a single Df
ht_plot_level_heights <- merge(mean_height_t1, mean_height_t2, by="PLT_CN2")

```

## Estimating net change in tree height using the Horvitz-Thompson estimator

```{r}
# Number of remeasurement plots after all filters
fia_n <- nrow(ht_plot_level_heights)

# Estimating mean tree height at t1
t1_ht <- horvitzThompson(y = ht_plot_level_heights$`Time 1`, 
                         N = fia_n,
                         var_est = TRUE,
                         var_method = "LinHH")


# Estimating mean tree height at t2
t2_ht <- horvitzThompson(y = ht_plot_level_heights$`Time 2`,
                         N = fia_n,
                         var_est = TRUE,
                         var_method = "LinHH")

# Calculating difference of t2 and t1 means as proxy for net growth 
nfi_net_growth <- t2_ht$pop_mean - t1_ht$pop_mean
nfi_net_growth

```
## Confidence Interval of NFI data
```{r}
# nfi_t1_var <- var(ht_plot_level_heights$`Time 1`)
# nfi_t2_var <- var(ht_plot_level_heights$`Time 2`)
nfi_t1_var <- t1_ht$pop_mean_var
nfi_t2_var <- t2_ht$pop_mean_var
nfi_cov <- cov(ht_plot_level_heights$`Time 1`, ht_plot_level_heights$`Time 2`)
nfi_pooled_var <- (nfi_t1_var + nfi_t2_var) - nfi_cov
ci_nft <- 1.96 * sqrt(nfi_pooled_var) 
ci_nft
```

## Pivoting
```{r}
ht_plot_level_heights_longer <- ht_plot_level_heights |> 
  pivot_longer(
    cols = c(`Time 1`, `Time 2`),
    names_to = "Variable",
    values_to = "Value"
  )
ht_plot_level_heights_longer
```

## Graphing
```{r}
nfi_t1_heights_plot <- ggplot(ht_plot_level_heights, aes(x = `Time 1`)) + 
  geom_histogram(binwidth = 1, fill = "#3689D2", color = "black") +
  scale_x_continuous(breaks = seq(0, 30, by = 5)) + 
  geom_vline(xintercept=mean(ht_plot_level_heights$`Time 1`), color='red', lwd=.75)+
  annotate("text", x = 12, y = 25, size=7, face='bold', color='black', label = glue("Mean: {round(mean(ht_plot_level_heights$`Time 1`), 2)}"))+
  labs(title = "Previous",
       x = "Height (m)",
       y = "Frequency")+
  theme_minimal() +
  customTheme

nfi_t1_heights_plot

nfi_t2_heights_plot <- ggplot(ht_plot_level_heights, aes(x = `Time 2`)) + 
  geom_histogram(binwidth = 1, fill = "#8A7D63", color = "black") +
  scale_x_continuous(breaks = seq(0, 30, by = 5)) + 
  geom_vline(xintercept=mean(ht_plot_level_heights$`Time 2`), color='red', lwd=.75)+
  annotate("text", x = 13.75, y = 25, size=7, face='bold', color='black', label = glue("Mean: {round(mean(ht_plot_level_heights$`Time 2`), 2)}"))+
  labs(title = "Current",
       x = "Height (m)",
       y = "Frequency")+
  theme_minimal() +
  customTheme

nfi_t2_heights_plot

combined_nfi_heights_plot <- (nfi_t1_heights_plot + nfi_t2_heights_plot)

ggsave(r"(C:\Users\poncy\Desktop\Masters\Research\Publication\figures\combined_nfi_heights_plot.png)", plot = combined_nfi_heights_plot, bg = "transparent",
width = 16, height = 8, units = "in", dpi = 300)

combined_nfi_heights_plot
```

